Supplementary Information: \nTree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest

### 2020-07-10 manuscript/response_to_reviews.Rmd

[*issue #94*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/94)

I suggest to show some relationship, e.g. between Rt and traits in figures. The study only presents two figures and there is plenty of room to present results more clearly with extra figures.

[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)

[*issue #100*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)

@@

[*issue #94*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/94)

**We have gone through the manuscript / tables and figures and updated where it said "dAICc" to be $\Delta$AICc. **

Following the two previous comments, I suggest to leave in Table 5 just the best model (rather that a multimodel inference as is presented now) to discuss more clearly the covariates included. And I suggest discussing only the model with all years together. This would be more robust and avoid ramble differences between years that mostly look spurious and with a difficult physiological explanation. This would help to ensure a better use of the Resilience indices.

*How about this: *

*(1) keep a version of table 5 in the main text, showing just the top model for Rt, and maybe also that of $Rt_{ARIMA}$.  *

*(2) put full versions of table 2 (for Rt and $Rt_{ARIMA}$), including all models within the top dAICc=2, in the appendix*

@@

related to [*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)

related to [*issue #99*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/99)

**We do want to show the three separate drought periods, as we feel that this makes the analysis richer and helps to separate real drivers from spurious correlations. However, in order to make models more consistent across years and reduce spurious correlations, we now include the criteria that for a trait to be included as a candidate variable in the full model, the sign of coefficients must be consistent across all droughts (in addition to meeting the AIC criteria, as before). This resulted in only two candidate variables for the main model ($PLA_{dry}$ and $\pi_{tlp}$). (Removed was ring porosity, which had opposite effects in different droughts.)**

@@

The authors might like to consider including Table 4 as a supplement, because as it stands it looks like an exploratory analysis. If the ‘single-variable approach’ is not just exploratory, state it clearly and why it is needed besides the full models with several parameters.

*This table is definitely not just exploratory, and is the source for supporting/ rejecting predictions in Table 1. Thus, keeping it in the main text could be justified. That said, it presents mainly negative results, and I'd like to reduce the number of tables, so moving it to the appedix would be fine.*

@@

[*issue #95*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/95)

related to [*issue #99*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/99)

@@

Importantly, the drought selection would be biased if years were selected as from line 202, which is a common mistake in similar studies (the original of Lloret et al. included). Droughts must be selected based only on some climatic/drought index (PDSI, SPEI,…) and independently of growth. Otherwise the analysis is biased towards years with low growth, and different questions would arise with a difficult answer. If the authors select only those droughts where there is a growth decrease: is that decrease evident in all species and similar across species? What happens with more intense droughts that those selected (if any), why they do not force a decrease in growth if drought is a main forcing agent? I see that 1966 and 1999 are the driest (as in PDSI) in Figure 1, but there is at least one year around 1955 that it is as dry or drier than 1977. In line 204 it is stated that they are the driest. If this is the case, the analysis should be sound but it should be deleted what is stated in lines 202-203 where authors say that they selected years also based on low growth. In fact, that should be a result after having selected the drought years only based on dryness (as in lines 208-209). Line 211 then they search for pointer years… a little bit messy. Please, explain better, and it should follow: 1) selection of dry years without regarding growth data; 2) characterise growth those dry years; 3) If the authors want to compare with other pointer years, fine, but that should be a third independent step properly justified and explained. Following this rationale, then in line 221 it should be difficult that Rt>1, had we selected those years with a negative growth anomaly. Please, explain clearly and consistently.

**We have made the suggested change: we now present only the PDSI criteria in the methods, and mention the fact that the selected years were also the ones with largest growth reductions during the study period in the results section. Using just the PDSI criterion does not change the years analyzed.**

@@

Other minor comments:

- Line 113: correct ‘positively’.

**done**

- Line 134: cambial growth also includes that of phloem; so better refer to ‘xylem growth’.

**done**

- Line 161: how many species were estimated for height like this? Wouldn’t it be possible for authors to go to the field and perform some height estimations? This 
should be rather straightforward and the authors would increase accuracy across species in height estimation.

**We have specified how many species used the interspecific allometry (n=2, JUNI and FRAM). Because of COVID-19, it would not be possible to go collect more data for these species.**

- Lines 166-167: please explain further the hypothesis and why.

**We have reworded this sentence to read, "While some tree crowns undoubtedly changed position over the past several decades, in this case the bias would be unlikely to result in false acceptance of the prediction that dominant trees have the lowest $Rt$ (*i.e.*, type I error unlikely, type II error possible), making our hypothesis test conservative."**

- Line 222: what about climate in years before 1991?

**We no longer mention 1991, as it does not meet the PDSI criteria.**

- Line 231: define AICc.

**This is done (with added reference)**

- Line 234-247: it is not clear what the authors did, please explain clearer this paragraph and the next and why you did two different (yet similar) analyses (single-variable and full model).

**This section has been re-written.**

@@

- Line 241-242: enough to say once that models were selected based on AIC (line 237-238) and then one criterion, why dAIC<1 here and dAIC<2 before? The classic assumption is 2 (e.g. Burnham & Anderson 2004). See also my other comment above.

**We have modified this text to read as follows:**

**"Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to a corresponding null model lacking the trait improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table 4). We note that the $\Delta$AICc $\ge$ 1.0 criterion is not a test of significance, but of whether the variable has enough influence to be considered as a *candidate* variable in full models."**

@@

- Line 255-256: but this was already an assumption in M-M (line 202-203). Please, see my comment below and be consistent along the text.

**As stated above, we have removed the $Rt$ criterion from the drought identification process, and now present this information for the first time in this section, adding the following statement: "Across the entire study period (1950-2009), the focal drought years were the three years with the largest fraction of trees exhibiting $Rt \le 0.7$."**

@@

- Line 258: so this means that they are no sensitive to the drought? Which trees, any species in particular? Is this a consequence of stand dynamics shadowing the impact of drought events?

**It's not unprecedented to see some trees exhibit increased growth during drought, and not surprising. Stand dynamics and other factors result in a lot of scatter in the relationship between tree growth and climate, so while the distribution of $Rt$ in Fig. 1b was shifted left (normal year should be normally distributed around 1), there still are some trees with increased growth. Our analysis helps to work out which trees these were (often shorter trees).**

**We address this question in the discussion, where we have added a statement about how this is not unprecedented: "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. 1b), potentially due to decreased leaf area of competitors during the drought, and consistent with prior observations that smaller trees can exhibit increased growth rates during drought (Bennett et al. 2015). It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation"**

@@

- Line 266-267: why only in 1966? I think this shows that this variable is not properly assessed. The older the date, the more accurate the estimation of crown position and height were.

**It is true that crown position estimates become more accurate in later dates, but height should be equally reliable across years, having been estimated based on reconstructed DBH. This was not previously clearly specified in the methods. To clarify, we have added the following statement: "We then used these allometries to estimate $H$ for each drought year, $Y$, based on reconstructed $DBH_Y$."**

**We agree that the assessment of crown position is somewhat problematic, both because of its increasing inaccuracy going back through time, and because of collinearity with $H$. However, given that it is both biologically interesting and comes out in our top models, we have chosen to include it. We note that this makes the test of our hypothesis conservative. We are transparent about this issue, to which we devote a paragraph in the discussion (par. 3).**

@@

- Line 275-280: those profiles were assumed a priori in the introduction and are common to any forest. They are indeed interesting to characterise the forest, but they do not add to individual tree variability. You could think to send Figure 2 to an appendix, in consequence, and use figures to show more specific results related to your individual tree analyses.

*Regarding Fig. 2, my current inclination is to keep it While I think its true that such profiles are common, I'm not currently aware of a lot of studies showing such profiles. I just prepared a grant proposal related to this, and current intern Nidhi Vinod is working on a review on vertical profiles in temperature, so I'd hope I'd be relatively up to speed... I'll check into this more and come back to it.*

@@

[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)

- Line 330-331: any evidence to support this? It looks rather speculative. As suggested above, does any species particularly show positive Rt or positive Rt are randomly distributed in your trees?

*Current text, with phrase of concern in bold: "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. 1b), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]." *

- Line 331-333: not so sure about this. It can also mean that there is no causal relationship between Rt and covariates, mostly because radial growth Rt should not be used as an integral proxy of tree resistance to drought, such as it is often used now in the literature.

*Tree-ring data inherently have a lot of variation. Neil could probably write a better response here, with refs.*

- Line 334-339: indeed, assessment of these two variables is not conclusive, as suggested above (just data from 2018, from the past estimated and correlated).

**Please see response above.**

- Line 348: non-drought.

**Fixed.**

- Line 349: do authors really think that lower radial growth as from Rt on a specific year means that the tree is more vulnerable? I think this is over conclusive.

**We did not intend the meaning here as vulnerability *to mortality* (assuming that's the objection here). We have reworded this sentence to read, "Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019]."**

- Line 400: I do not agree we can directly link Rt with tree vulnerability to drought; we might be currently overinterpreting these resilience indices extrapolating to physiological conclusions that do not necessary need to hold.

**See above comment. We have reworded this sentence to read, "In the meantime, the results of this study advance our knowledge of the factors conferring drought resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought."**

Table 1: do ‘-‘ signs in the table correspond to ‘no’ (non-significant)? Please explain.

**This is now clarified in the table footnote.**

Table 3: please, add also the number of trees, it is more informative than the number of cores to check sample size in your study.

**In this study, we have one core per tree, now specified in the methods. We have changed the table header to "n trees".**

Table 3: the authors could show some test to see if the traits presented are different among species.

**maybe include these in fig with $Rt$ across species? [*issue #100*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)*

Table S1 and S2: correct R2.

*IM I added the correct Latex notation but it's not rendering it when I knit the file... (KAT: I just tried and wasn't able to get it working.*

Table S3: please, explain what ‘rank’ means.

**Done.**

Figure S3: I understand that authors only have height data from 2018 and height for the other dates looks trivial: indeed there will be an increase in height with an increase in DBH in time. The question is whether there was any stand dynamics involved in those height series that had to be included in the models.

**We recognize the importance of stand dynamics, and have added specific mention to the discussion: "It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation,... "**

### 2020-07-10 manuscript/climate_sensitivity_SI.Rmd
worked on reviewer responses
all of R1 (although upen issues remain)

#97

bark <- read.csv("tables_figures/publication/tableS1_bark_regression.csv", stringsAsFactors = FALSE)
bark[,2] <- NULL
colnames(bark) <- c("Species", "Equations", "$R^{2}$")
kable(bark, booktabs=TRUE, format = "latex", caption = "Table S1: Species-specific bark thickness regression equations") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

\newpage
```{r Table S2, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)

heights <- read.csv("tables_figures/publication/tableS2_height_regression.csv", stringsAsFactors = FALSE)
heights[,2] <- NULL
colnames(heights) <- c("Species", "Equations", "$R^{2}$")
kable(heights, booktabs=TRUE, format = "latex", caption = "Table S2: Species-specific height regression equations") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

\newpage
```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
tableS3 <- read.csv("tables_figures/publication/tables3_drought_stats.csv", stringsAsFactors = FALSE)
kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts. ") %>%
  pack_rows("focal droughts", 1, 4, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
  pack_rows("", 5, 8, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 9, 12, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("other", 13, 16, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```
Rank refers to ...**XX**

>>>>

bark <- read.csv("tables_figures/publication/tableS1_bark_regression.csv", stringsAsFactors = FALSE)
bark[,2] <- NULL
colnames(bark) <- c("Species", "Equations", "$R^2$")
kable(bark, booktabs=TRUE, format = "latex", caption = "Table S1: Species-specific bark thickness regression equations") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

\newpage
```{r Table S2, eval=TRUE, echo=FALSE,  warning=FALSE}
library(knitr)
library(kableExtra)

heights <- read.csv("tables_figures/publication/tableS2_height_regression.csv", stringsAsFactors = FALSE)
kable(heights, booktabs=TRUE, format = 'latex', caption = "Table S2: Species-specific height regression equations", escape = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position", protect_latex = TRUE)
```

\newpage
```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
tableS3 <- read.csv("tables_figures/publication/tables3_drought_stats.csv", stringsAsFactors = FALSE)
kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Monthly Palmer Drought Severity Index (PDSI), and its rank among all years between 1950 and 2009 (driest=1), for focal droughts. ") %>%
  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
  pack_rows("", 5, 8, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 9, 12, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position", protect_latex = TRUE)
```
